---
title: "Loading in Data"
---

This week we aim to supplement the assigned readings and homework with a few blurbs and examples in `R`. We will learn how to install packages and load data into the `RStudio` environment, and will discuss how to build your own functions in `R`.

Two of the first skills that you are going to need to learn are 1) how to install / load your required packages into the `R` environment and 2) how to import data from a file. More often than not, these will be the first two steps of your analysis (though if you are working on more complex projects you may find it easier to keep track of things if you load import your data and packages as needed).

Note that much, but not all of what we discuss here is covered in two videos available to you on the [GradStats Youtube Channel](https://youtube.com/playlist?list=PLLv6V3KCs8wMoc5i2YT1YCmZudgne7Xzj).

# Loading in data

Typically the file types that are used by beginners in `R` are plain text and delimited. They may have the extension "txt", "csv", or "dat" for example. These may become more sophisticated you progress, for example you can load in proprietary types like SPSS and STATA, but for this course we will mostly use plain text files (although later in the course I will show you how to load in Excel files and SPSS files).

## Loading in local data using a GUI:

As with loading packages, you can also load in a file containing data using the RStudio GUI. See [this video](https://youtu.be/NqvxpK495iU).

Again, this is only preferred if you are not sharing your analysis. If you are sharing your analysis (as in this class) you need to do the command line. Fortunately, RStudio creates the appropriate command-line for you to copy and paste. For example, loading one of your homework data sets can be accomplished by:

```{r}
#| eval: false
library(tidyverse)
LexicalDescisionData <- read_table("path/to/LexicalDescisionData.txt")
```

where `path/to/` is the path to the folder containing the file `LexicalDescisionData.txt`.

I've uploaded this file to OneDrive in the `walkthroughs and examples` folder. Please feel free to download the file and follow along is you wish. Note that *if you do elect to load in via the GUI you need to be sure to copy the output to your Rmd source. Otherwise, I will not be able to run it on my computer*.

## Importing data from the web

You might be saying to yourself, "but Tehran the entire reason you've got us learning `R` is for transparency and openness with our data. How would I be able to share in my code a data file that resides on my hard drive?!?""

Correct, you can't, but you can upload it to the internet and someone can access it from an online repository. Personally, I like to use [http://www.github.com](**Github**), but we'll save that for some advanced stuff later in the semester for those so inclined. Another alternative is to upload your entire folder to a project in [http://www.rstudio.cloud](**RStudio.cloud**).

If the data is uploaded directly from the web, we can load it in using the `read_delim()` function from above. This is some reaction time data taken from a website. Let's assign it to an object `RxData`



```{r}
#| eval: false
library(tidyverse) # no need to ask if already loaded
alcohol_use_data <- read_csv("https://tehrandav.is/psyc7014/practice_datasets/alcuseN6.csv", col_names = T)

```
```{r}
#| echo: false

library(tidyverse) # no need to ask if already loaded
alcohol_use_data <- read_csv("../practice_datasets/alcuseN6.csv", col_names = T)
```

To see what's going on with the additional calls in this line, run the following line to get help:

```{r}
#| eval: false
? read_csv
```

A document file should show up in your help tab, containing examples and describing what different arguments are doing. Search for `col_names` and try to figure out what's going on.

## Looking ahead...

This is probably a good place to stop for now. In the meantime try running the following 4 commands (assuming that you have imported alcohol_use_data) and think about what they are returning:

```{r}
#| eval: false

class(alcohol_use_data)
names(alcohol_use_data)
head(alcohol_use_data)
summary(alcohol_use_data)
```
